---
title: "Assisted Living"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
runtime: shiny
resource_files:
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.cpg
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.dbf
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.prj
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbn
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbx
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shp.xml
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)

```

```{r loading jon docs and shapefiles, cache=TRUE, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"),
         mif = ifelse(is.na(mif), "TC", mif),
         mif = as.factor(mif),
         mif = fct_relevel(mif, "NW", "NE", "WC", "EC", "SW", "SE", "TC"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)

```

<br>

# Summary

**Age 65+ per bed**

By far the highest ratios of population 65 and older per assisted living capacity are located in rural areas. In our entirely rural counties it's 33 people age 65 or older per assisted living capacity compared to 22, 20 and 15 in town/rural mix, urban/town/rural mix and entirely urban counties, respectively.

Counties on the west side of the state and in the central lakes region are particularly high.

**Assisted living capacity as percentage of nursing home bed declines**

To try and get a sense of the role assisted living might play in serving older populations we looked at the percentage this capacity makes up of all the declines in nursing home beds since 2005. 

Again, there are stark differences across rural and urban. As a state, assisted living capacity makes up nearly 500% of the losses in nursing home beds, but this is largely due to significantly higher density of them in urban areas. Entirely urban counties have assisted living capacity that makes up for over 800% of the loss in nursing homes. In entirely rural counties, the assisted living capacity does not fully make up for all the losses in nursing home beds - 93%. 

The percentages are particularly low in counties in the Northwest, Southwest, and Cass County.

<br>

# Data Prep

First, we will load the master file.

```{r load master file}
master.original <- read_csv("Data/Facilities/Master-provider-directory-2005-2024.csv") %>%
  mutate(Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

master.original %>%
  names() %>%
  kable()

master.age.65plus.county <- read_csv("Data/Population/Age/Master-age-65plus-county.csv") %>%
  mutate(Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

master.age.projections.65plus.county <- read_csv("Data/Population/Age/Master-age-projections-65plus-county.csv") %>%
  mutate(Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))
```

There are `r comma(ncol(master.original))` columns in the dataset and `r comma(nrow(master.original))` rows. Each row is a facility tied to a specific year.

The following table defines all of the columns.

```{r varaiable definition table}
field.outline <- read_excel("Data/Facilities/Field outline.xlsx") %>%
  as_tibble()

datatable(field.outline, class = "cell-border stripe", filter = "top", rownames = FALSE)

```

<br>

With this exploration, the primary data we want are only the assisted living facilities and their capacity. The column that's going to indicate this are the "ALL_PROV" and "ALL_CAPACITY" columns. 

Here is a list of all the facilities.

<br>

```{r assisted living fac}
master.al.fac <- master.original %>%
  filter(ALL_CAPACITY > 0) %>%
  select(year, HFID, NAME, COUNTY_NAME, LIC_TYPE, ALL_CAPACITY, countyfp, Dem_Desc, edr, planning.region, `MIF Region`)

datatable(master.al.fac, class = "cell-border stripe", filter = "top", rownames = FALSE,
          options = list(columnDefs = list(list(className = "dt-center", targets = 1:10)))) 
```

<br>

Unfortunately, they only started licensing these facilities in 2021. The law established regulatory standards governing the provision of housing and services in assisted living facilities and assisted living facilities with dementia care to help ensure the health, safety, well-being, and appropriate treatment of residents. 

In 2024, there are `r comma(nrow(master.al.fac))` facilities in Minnesota.

<br>

# Capacity

Since we only have one year of data, in this section we explore the differences in capacity across the state but no trends. To keep capacity relative, we will use a couple of difference ratios;

* ages 65+ per bed available, and
* ratio of skilled nursing home beds to assisted living beds.

<br>

## Age 65+ per bed{.tabset}

The charts below show the number of people that are 65 or older per assisted living bed. Consistently, rural areas have some of the highest number of 65+ per bed. In Minnesota, there are 17 people age 65+ per bed, however, it's significantly higher in rural Minnesota. In our entirely rural counties, there are 33 people age 65+ per assisted living bed, compared to 22 in town/rural mix, 20 in urban/town/rural mix, and 15 in entirely urban counties.

Counties along the borders and in the central lakes are particularly high.

<br>

```{r 65plus per bed data prep, echo=FALSE}
al.fac.county <- master.al.fac %>%
  group_by(year, countyfp) %>%
  summarize(ALL_CAPACITY = sum(ALL_CAPACITY)) %>%
  ungroup() %>%
  left_join(counties.regions, by = c("countyfp")) %>%
  left_join(master.age.65plus.county[,c(1,2,3)], by = c("countyfp", "year"))
```

### Minnesota

In Minnesota, there are 16.8 people age 65 or older per assisted living facility bed. 

<br>

```{r 65plus per bed mn}
al.fac.county %>%
  group_by(year) %>%
  summarize(ALL_CAPACITY = sum(ALL_CAPACITY),
            age.65plus = sum(age.65plus)) %>%
  ungroup() %>%
  mutate(age.65plus.per.bed = age.65plus / ALL_CAPACITY) 

```

<br>

### Planning Region

The seven county metro has significantly higher capacity than does other regions of Minnesota. The lowest capacity is Northwest MN where there are 23 people aged 65+ per assisted living bed. The remaining regions at 20 per bed except for Southeast where it is 18 per bed.

<br>

```{r 65plus per bed pr}
al.65plus.bed.pr <- al.fac.county %>%
  group_by(planning.region) %>%
  summarize(ALL_CAPACITY = sum(ALL_CAPACITY),
            age.65plus = sum(age.65plus)) %>%
  ungroup() %>%
  mutate(age.65plus.per.bed = age.65plus / ALL_CAPACITY,
         data_id = seq(n()))

al.65plus.bed.pr.plot <- ggplot(al.65plus.bed.pr, aes(planning.region, age.65plus.per.bed, fill = planning.region)) +
  geom_col_interactive(position = "dodge", aes(data_id = data_id, tooltip = paste(planning.region, "\nNumber of assisted living beds: ", comma(ALL_CAPACITY), "\nNumber of 65+: ", comma(ALL_CAPACITY), "\nAge 65+ per bed: ", comma(age.65plus.per.bed),  sep = ""))) +
  geom_label(aes(label = comma(age.65plus.per.bed, accuracy = 1)), show.legend = FALSE, color = "white", size = 5, position = position_dodge(width = .9)) +
  labs(x="", y = "", color="", title = "Number of people age 65+ per assisted living bed")+
  scale_y_continuous(labels=scales::comma)+
  theme_bar+
  scale_fill_manual(values = color.pr,
                    guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 18))


girafe(ggobj = al.65plus.bed.pr.plot, width_svg = 7, height_svg = 7) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE))      

```

<br>

### EDR

Again we can see significantly lower capacity in the most rural areas of Minnesota. The most rural regions in Northwest, Central and Southwest have significantly numbers of 65+ per bed. The highest is in EDR 6W where there are 28 people age 65+ per bed.

<br>

```{r 65plus per bed edr}
al.65plus.bed.edr <- al.fac.county %>%
  group_by(edr, planning.region) %>%
  summarize(ALL_CAPACITY = sum(ALL_CAPACITY),
            age.65plus = sum(age.65plus)) %>%
  ungroup() %>%
  mutate(age.65plus.per.bed = age.65plus / ALL_CAPACITY,
         data_id = seq(n()),
         edr = str_sub(edr, 1, 6),
         edr = trimws(edr, which = "both"))

al.65plus.bed.edr.plot <- ggplot(al.65plus.bed.edr, aes(edr, age.65plus.per.bed, fill = edr)) +
  facet_wrap(~planning.region, ncol = 2, scales = "free_x") +
  geom_col_interactive(position = "dodge", aes(data_id = data_id, tooltip = paste(edr, "\nNumber of assisted living beds: ", comma(ALL_CAPACITY), "\nNumber of 65+: ", comma(ALL_CAPACITY), "\nAge 65+ per bed: ", comma(age.65plus.per.bed),  sep = ""))) +
  geom_label(aes(label = comma(age.65plus.per.bed, accuracy = 1)), show.legend = FALSE, color = "white", size = 5, position = position_dodge(width = .9)) +
  labs(x="", y = "", color="", title = "Number of people age 65+ per assisted living bed")+
  scale_y_continuous(labels=scales::comma)+
  theme_bar+
  scale_fill_manual(values = color.edr,
                    guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 18))


girafe(ggobj = al.65plus.bed.edr.plot, width_svg = 7, height_svg = 10) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE))      

```

<br>

### RUCA

This chart really shows how the capacity changes depending on rurality. Our entirely rural counties have 33 age 65plus per bed, while town/rural mix has 22, urban/town/rural mix has 20, and entirely urban has 15.

<br>

```{r 65plus per bed ruca}
al.65plus.bed.ruca <- al.fac.county %>%
  group_by(Dem_Desc) %>%
  summarize(ALL_CAPACITY = sum(ALL_CAPACITY),
            age.65plus = sum(age.65plus)) %>%
  ungroup() %>%
  mutate(age.65plus.per.bed = age.65plus / ALL_CAPACITY,
         data_id = seq(n()))

al.65plus.bed.ruca.plot <- ggplot(al.65plus.bed.ruca, aes(Dem_Desc, age.65plus.per.bed, fill = Dem_Desc)) +
  geom_col_interactive(position = "dodge", aes(data_id = data_id, tooltip = paste(Dem_Desc, "\nNumber of assisted living beds: ", comma(ALL_CAPACITY), "\nNumber of 65+: ", comma(ALL_CAPACITY), "\nAge 65+ per bed: ", comma(age.65plus.per.bed),  sep = ""))) +
  geom_label(aes(label = comma(age.65plus.per.bed, accuracy = 1)), show.legend = FALSE, color = "white", size = 5, position = position_dodge(width = .9)) +
  labs(x="", y = "", color="", title = "Number of people age 65+ per assisted living bed")+
  scale_y_continuous(labels=scales::comma)+
  theme_bar+
  scale_fill_manual(values = color.ruca,
                    guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 18))


girafe(ggobj = al.65plus.bed.ruca.plot, width_svg = 7, height_svg = 7) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE))      

```

<br>

### County

The map shows that rural areas have significantly higher population of 65+ per assisted living bed.

<br>

```{r 65plus per bed county}
al.65plus.bed.county <- al.fac.county %>%
  mutate(age.65plus.per.bed = age.65plus / ALL_CAPACITY,
         age.65plus.per.bed.bins = cut(age.65plus.per.bed,
                                       breaks = c(0, 10, 20, 30, 40, 1000),
                                       labels = c("9 to 10", "10 to 20", "20 to 30", "30 to 40", "40+"))) %>%
  right_join(mn_counties[,c(5,7)], by = "countyfp")

names(al.65plus.bed.county)

al.65plus.bed.county.plot <- ggplot(al.65plus.bed.county) +
  geom_sf_interactive(color = "grey85", aes(geometry = geometry, fill = age.65plus.per.bed.bins, data_id = countyfp, tooltip = paste(Name, "\nYear: ", year, "\nPopulation age 65+: ", comma(age.65plus), "\nNumber of assisted living beds: ", comma(ALL_CAPACITY), "\nNumber age 65+ per bed: ", comma(age.65plus.per.bed), sep = ""))) +
  theme_sf+
  scale_fill_manual(values = brewer.pal(n = 6, "PuBu")) +
  labs(title = "Population 65+ per assisted living bed") +
  theme(legend.position = "bottom",
        text = element_text(size = 18))

girafe(ggobj = al.65plus.bed.county.plot, width_svg = 7, height_svg = 7) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE))     

```

<br>

## Assisted living capacity as perentage of nursing home decline{.tabset}

Let's see how much these beds make up for the loss in nursing home beds.

The charts below is an attempt to tell the story of how much assisted living capacity can make up for the decline in nursing home beds. Even though that isn't an apples-to-apples comparison, it does at least provide an idea of the role that assisted living is attempting to fill.

In Minnesota, assisted living makes up almost 500% of the decline in nursing home beds statewide. 

At a regional level, the percentage varies significantly depending on rural-urbanness. The assisted living capacity in the seven county metro and Central planning regions make up a significantly higher percentage of their declines in nursing home beds - 795% and 622% respectively. However, it drops pretty significantly in the other regions. Northwest and Southwest in particular only have a bit over 200% of their decline in nursing home beds made up through assisted living.

When grouping counties by rural-ness, entirely rural counties don't even have enough assisted living capacity to make up for the entire decline of nursing home beds - 93%. The county map shows that the lowest percentages are located in Northwest and Southwest Minnesota, as well as Cass County in North Central.

<br>

```{r prep al as pct of snf bed loss since 2005, echo=FALSE}
master.snf <- master.original %>%
  select(year, LIC_TYPE, NH_BEDS, SNF_BEDS, SNFNF_BEDS, NF1_BEDS, NF2_BEDS, HFID, NAME, CITY, STATE, ZIP, COUNTY_NAME, countyfp, Dem_Desc, edr, planning.region)

nh.nf2.beds.master <- master.snf %>%
  mutate(nh.nf2.beds = NH_BEDS + NF2_BEDS) %>%
  filter(nh.nf2.beds > 0)

snf.bed.change.al <- nh.nf2.beds.master %>%
  filter(year %in% c(min(nh.nf2.beds.master$year), max(nh.nf2.beds.master$year))) %>%
  group_by(year, countyfp) %>%
  summarize(nh.nf2.beds = sum(nh.nf2.beds)) %>%
  ungroup() %>%
  pivot_wider(names_from = year, values_from = nh.nf2.beds) %>%
  mutate(snf.bed.change = `2005` - `2024`) %>%
  rename(snf.bed.2005 = `2005`,
         snf.bed.2024 = `2024`) %>%
  left_join(al.fac.county, by = "countyfp") %>%
  select(-year)

```

### Minnesota

The assisted living capacity in Minnesota is nearly 500% of the decline in nursing home beds.

<br>

```{r al as percent of snf bed change mn, echo=FALSE}
al.pct.snf.change.mn <- snf.bed.change.al %>%
  summarize(snf.bed.2005 = sum(snf.bed.2005, na.rm = TRUE),
            snf.bed.2024 = sum(snf.bed.2024, na.rm = TRUE),
            ALL_CAPACITY = sum(ALL_CAPACITY, na.rm = TRUE)) %>%
  mutate(bed.change = snf.bed.2024 - snf.bed.2005,
         al.pct.snf.change = (ALL_CAPACITY / bed.change) * -1)

kable(format = "html", al.pct.snf.change.mn, escape = F) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE) %>%
  scroll_box(width = "100%", height = "300px")

```

<br>

### Planning Region

Assisted living units in more metropolitan areas make up a far larger percentage of the decline in nursing home beds. Assisted living facilities in the seven county metro have 800% more capacity than the decline in nursing home beds. In our rural areas, it averages 200% to 400%.

<br>

```{r al as percent of snf bed change pr, echo=FALSE}
al.pct.snf.change.pr <- snf.bed.change.al %>%
  filter(!is.na(planning.region)) %>%
  group_by(planning.region) %>%
  summarize(snf.bed.2005 = sum(snf.bed.2005, na.rm = TRUE),
            snf.bed.2024 = sum(snf.bed.2024, na.rm = TRUE),
            ALL_CAPACITY = sum(ALL_CAPACITY, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(bed.change = snf.bed.2024 - snf.bed.2005,
         al.pct.snf.change = (ALL_CAPACITY / bed.change) * -1,
         data_id = seq(n()))

al.pct.snf.change.pr.plot <- ggplot(al.pct.snf.change.pr, aes(planning.region, al.pct.snf.change, fill = planning.region)) +
  geom_col_interactive(position = "dodge", aes(data_id = data_id, tooltip = paste(planning.region, "\nSNF beds in 2005: ", comma(snf.bed.2005), "\nSNF beds in 2024: ", comma(snf.bed.2024), "\nChange in SNF beds: ", comma(bed.change), "\nAssisted living capacity in 2024: ", comma(ALL_CAPACITY), "\nAssisted living capacity as percentage of SNF bed change: ", percent(al.pct.snf.change, accuracy = .1), sep = ""))) +
  geom_label(aes(label = percent(al.pct.snf.change, accuracy = .1)), show.legend = FALSE, color = "white", size = 5, position = position_dodge(width = .9)) +
  labs(x="", y = "", color="", title = "Assisted living capacity as a percentage of\nnursing home bed declines")+
  scale_y_continuous(labels=scales::percent)+
  theme_bar+
  scale_fill_manual(values = color.pr,
                    guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = al.pct.snf.change.pr.plot, width_svg = 7, height_svg = 7) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE))      

```

<br>

### EDR

EDR 7W has an insane percentage of assisted living capacity to the decline in SNF beds at over 1100%. Other rural regions are significantly lower - they are around 100 to 350%.

<br>

```{r al as percent of snf bed change edr, echo=FALSE}
al.pct.snf.change.edr <- snf.bed.change.al %>%
  filter(!is.na(planning.region)) %>%
  group_by(edr, planning.region) %>%
  summarize(snf.bed.2005 = sum(snf.bed.2005, na.rm = TRUE),
            snf.bed.2024 = sum(snf.bed.2024, na.rm = TRUE),
            ALL_CAPACITY = sum(ALL_CAPACITY, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(bed.change = snf.bed.2024 - snf.bed.2005,
         al.pct.snf.change = (ALL_CAPACITY / bed.change) * -1,
         data_id = seq(n()),
         edr = str_sub(edr, 1, 6))

al.pct.snf.change.edr.plot <- ggplot(al.pct.snf.change.edr, aes(edr, al.pct.snf.change, fill = edr)) +
  facet_wrap(~planning.region, ncol = 2, scales = "free_x") +
  geom_col_interactive(position = "dodge", aes(data_id = data_id, tooltip = paste(edr, "\nSNF beds in 2005: ", comma(snf.bed.2005), "\nSNF beds in 2024: ", comma(snf.bed.2024), "\nChange in SNF beds: ", comma(bed.change), "\nAssisted living capacity in 2024: ", comma(ALL_CAPACITY), "\nAssisted living capacity as percentage of SNF bed change: ", percent(al.pct.snf.change, accuracy = .1), sep = ""))) +
  geom_label(aes(label = percent(al.pct.snf.change, accuracy = .1)), show.legend = FALSE, color = "white", size = 5, position = position_dodge(width = .9)) +
  labs(x="", y = "", color="", title = "Assisted living capacity as a percentage of\nnursing home bed declines")+
  scale_y_continuous(labels=scales::percent)+
  theme_bar+
  scale_fill_manual(values = color.edr,
                    guide = guide_legend(ncol = 3)) +
  theme(legend.position = "none",
        text = element_text(size = 18))


girafe(ggobj = al.pct.snf.change.edr.plot, width_svg = 7, height_svg = 7) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE))      

```

<br>

### RUCA

This shows pretty clearly how stark the difference is in assisted living capacity making up the decline in nursing home beds. Enrirely rural counties have assisted living capacity that doesn't make up for the loss in nursing home beds. 

<br>

```{r al as percent of snf bed change ruca, echo=FALSE}
al.pct.snf.change.ruca <- snf.bed.change.al %>%
  filter(!is.na(planning.region)) %>%
  group_by(Dem_Desc) %>%
  summarize(snf.bed.2005 = sum(snf.bed.2005, na.rm = TRUE),
            snf.bed.2024 = sum(snf.bed.2024, na.rm = TRUE),
            ALL_CAPACITY = sum(ALL_CAPACITY, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(bed.change = snf.bed.2024 - snf.bed.2005,
         al.pct.snf.change = (ALL_CAPACITY / bed.change) * -1,
         data_id = seq(n()))

al.pct.snf.change.ruca.plot <- ggplot(al.pct.snf.change.ruca, aes(Dem_Desc, al.pct.snf.change, fill = Dem_Desc)) +
  geom_col_interactive(position = "dodge", aes(data_id = data_id, tooltip = paste(Dem_Desc, "\nSNF beds in 2005: ", comma(snf.bed.2005), "\nSNF beds in 2024: ", comma(snf.bed.2024), "\nChange in SNF beds: ", comma(bed.change), "\nAssisted living capacity in 2024: ", comma(ALL_CAPACITY), "\nAssisted living capacity as percentage of SNF bed change: ", percent(al.pct.snf.change, accuracy = .1), sep = ""))) +
  geom_label(aes(label = percent(al.pct.snf.change, accuracy = .1)), show.legend = FALSE, color = "white", size = 5, position = position_dodge(width = .9)) +
  labs(x="", y = "", color="", title = "Assisted living capacity as a percentage of\nnursing home bed declines")+
  scale_y_continuous(labels=scales::percent)+
  theme_bar+
  scale_fill_manual(values = color.ruca,
                    guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))


girafe(ggobj = al.pct.snf.change.ruca.plot, width_svg = 7, height_svg = 7) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE))      

```

<br>

### County

The county map shows that some of the lowest percentage of assisted living capacity making up for nursing home declines are located in Northwest and Southwest Minnesota.

<br>

```{r al as percent of snf bed change county, echo=FALSE}
al.pct.snf.change.county <- snf.bed.change.al %>%
  filter(!is.na(planning.region)) %>%
  filter(snf.bed.change > 0) %>%
  mutate(al.pct.snf.change = ALL_CAPACITY / snf.bed.change,
         al.pct.snf.change.bins = cut(al.pct.snf.change,
                                      breaks = c(0, 1, 2, 3, 4, 100000),
                                      labels = c("Less than 100%", "100% to 200%", "200% to 300%", "300% to 400%", "More than 400%"))) %>%
  right_join(mn_counties[,c(5,7)], by = "countyfp")

al.pct.snf.change.county.map <- ggplot(al.pct.snf.change.county) +
  geom_sf_interactive(color = "grey85", aes(geometry = geometry, fill = al.pct.snf.change.bins, data_id = countyfp, tooltip = paste(Name, "\nSNF beds in 2005: ", comma(snf.bed.2005), "\nSNF beds in 2024: ", comma(snf.bed.2024), "\nChange in SNF beds: ", comma(snf.bed.change), "\nAssisted living capacity in 2024: ", comma(ALL_CAPACITY), "\nAssisted living capacity as percentage of SNF bed change: ", percent(al.pct.snf.change, accuracy = .1), sep = ""))) +
  theme_sf+
  scale_fill_manual(values = brewer.pal(n = 6, "PuBu")) +
  labs(title = "Assisted living capacity as a percentage of\nnursing home bed declines") +
  theme(legend.position = "bottom",
        text = element_text(size = 18))


girafe(ggobj = al.pct.snf.change.county.map, width_svg = 7, height_svg = 7) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_sizing(rescale = FALSE))      

```

<br>