---
title: "Employment"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
runtime: shiny
resource_files:
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.cpg
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.dbf
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.prj
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbn
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbx
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shp.xml
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
```

```{r loading jon docs and shapefiles, cache=TRUE, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         mif = ifelse(is.na(mif), "TC", mif),
         mif = as.factor(mif),
         mif = fct_relevel(mif, "NW", "NE", "WC", "EC", "SW", "SE", "TC"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#654C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4565b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.edr.simple <- c("EDR 1" = "#b3cde3", "EDR 2" = "#8c96c6", "EDR 3" = "#fe9929", "EDR 4" = "#8856a7", "EDR 5" = "#810f7c", "EDR 6E" = "#e5f5f9", "EDR 6W" = "#bdc9e1", "EDR 7E" = "#99d8c9", "EDR 7W" = "#2ca25f", "EDR 8" = "#74a9cf", "EDR 9" = "#0570b0", "EDR 10" = "#d7301f", "EDR 11" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4565b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)

```

```{r master files, echo=FALSE}
job.vacancies.original <- read_csv("Data/Job vacancies/Master-job-vacancies-home-health-nursing-pr.csv") %>%
  mutate(planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"))

master.jv.pr <- job.vacancies.original %>%
  mutate(data_id = seq(n()))

oes.mn.original <- read_csv("Data/Employment/OES_MN.csv")

master.oes.mn <- oes.mn.original

```

<br>

# Summary

Stuff

<br>

# Job vacancies{.tabset}

Let's take a look at the job vacancy rate for these.

**Home Health Aides**

There is consistently hundreds of vacancies in every planning region since 2005. A large increase occurred after 2018 and by 2020 and through the pandemic the vacancies exploded into nearly or over a thousand positions. There has been a bit of a recovery since the pandemic for many regions except Southeast.

<br>

**Registered Nurses**

There has been consistent growth in openings for registered nurses since 2005. There was a huge spike during the pandemic but a bit of recovery since than. But even the pre-pandemic levels were severe.

```{r job vacancies index pr, echo=FALSE}
jv.index.pr <- master.jv.pr %>%
  group_by(planning.region, new.title) %>%
  mutate(jv.index = (vacancies - vacancies[year == min(year)]) / vacancies[year == min(year)]) %>%
  ungroup() %>%
  mutate(new.title = ifelse(new.title == "Registered Nurses", "Registered Nurses", "Home Health Aides"),
         data_id = seq(n()))
```

<br>

## Home Health Aides

<br>

```{r jv rate home health aides pr, echo=FALSE}
jv.index.home.health.aides.pr.plot <- ggplot(data = filter(jv.index.pr, new.title == "Home Health Aides"), aes(year, vacancies, color = planning.region)) +
  facet_wrap(~planning.region, ncol = 2, scales = "free_y") +
  geom_hline(yintercept = 0, color = "black") +
  geom_line(se = FALSE, size = 3) +
  geom_point_interactive(size = 2.5, aes(data_id = data_id, tooltip = paste(planning.region, "\nYear: ", year, "\nOccupation: ", new.title, "\nNumber of job vacancies: ", comma(vacancies), sep = ""))) +
  labs(x="", y = "", color="", title = "Home Health Aides - Vacancies")+
  geom_label_repel(data = filter(jv.index.pr, year == max(year), new.title == "Home Health Aides"), aes(label = comma(vacancies, accuracy = 1)), show.legend = FALSE, size = 5) +
  scale_y_continuous(labels=scales::comma)+
  scale_x_continuous(breaks = seq(1900, 2050, 5)) +
  theme_line+
  scale_color_manual(values= color.pr,
                     guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))

girafe(ggobj = jv.index.home.health.aides.pr.plot, width_svg = 7, height_svg = 7) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE),
                 opts_sizing(rescale = FALSE))


```

<br>

## Registered Nurses

<br>

```{r jv rate registered nurses pr, echo=FALSE}
jv.index.reg.nurses.pr.plot <- ggplot(data = filter(jv.index.pr, new.title == "Registered Nurses"), aes(year, vacancies, color = planning.region)) +
  facet_wrap(~planning.region, ncol = 2, scales = "free_y") +
  geom_hline(yintercept = 0, color = "black") +
  geom_line(se = FALSE, size = 3) +
  geom_point_interactive(size = 2.5, aes(data_id = data_id, tooltip = paste(planning.region, "\nYear: ", year, "\nOccupation: ", new.title, "\nNumber of job vacancies: ", comma(vacancies), sep = ""))) +
  labs(x="", y = "", color="", title = "Registered Nurses - vacancies")+
  geom_label_repel(data = filter(jv.index.pr, year == max(year), new.title == "Registered Nurses"), aes(label = comma(vacancies, accuracy = 1)), show.legend = FALSE, size = 5) +
  scale_y_continuous(labels=scales::comma)+
  scale_x_continuous(breaks = seq(1900, 2050, 5)) +
  theme_line+
  scale_color_manual(values= color.pr,
                     guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 25, hjust = .9))

girafe(ggobj = jv.index.reg.nurses.pr.plot, width_svg = 7, height_svg = 7) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE),
                 opts_sizing(rescale = FALSE))


```


<br>

# Vacancies and Employment

So the big question is whether this is growth in the labor force - essentialyl are the vacancies on top of employment, or are the vacancies due to people leaving the position. 

<br>

```{r oes home health and registered nursing prep, echo=FALSE}
oes.home.health.reg.nurse.mn <- master.oes.mn %>%
  filter(occ_code %in% c("29-1111", "31-1000", "399021", "29-1141", "31-1100"))

names(oes.home.health.reg.nurse.mn)
```