---
title: "HCBS"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
runtime: shiny
resource_files:
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.cpg
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.dbf
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.prj
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbn
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbx
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shp.xml
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)

```

```{r loading jon docs and shapefiles, cache=TRUE, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", linewidth  = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", linewidth = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial") ,
        plot.title.position = "plot",
        plot.title = element_text(face = "bold"))

regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"),
         mif = ifelse(is.na(mif), "TC", mif),
         mif = as.factor(mif),
         mif = fct_relevel(mif, "NW", "NE", "WC", "EC", "SW", "SE", "TC"))


color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "grey", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("/Users/kellyasche/Library/CloudStorage/GoogleDrive-kasche@ruralmn.org/My Drive/Data Prep/R Projects/Shapefiles/County shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)

```

```{r master hcbs data}
hcbs.qcew.pr.original <- read_csv("Data/QCEW/QCEWResults.csv")

master.hcbs.pr.qcew <- hcbs.qcew.pr.original %>%
  select(periodyear, naicstitle, areaname, empYear, avg_estabYear, totwageYear, avgwkwageYear) %>%
  rename(year = periodyear,
         planning.region = areaname) %>%
  mutate(planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = str_replace(planning.region, ", MN", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         planning.region = trimws(planning.region, which = "both")) %>%
  complete(planning.region, year) %>%
  filter(year != max(year)) %>%
  mutate(avg_estabYear = ifelse(is.na(avg_estabYear), 10, avg_estabYear),
         empYear = ifelse(is.na(empYear), 376, empYear),
         totwageYear = ifelse(is.na(totwageYear), 5712696, totwageYear),
         avgwkwageYear = ifelse(is.na(avgwkwageYear), 292.18, avgwkwageYear))

hcbs.qcew.mn.original <- read_csv("Data/QCEW/QCEWResultsmn.csv")

master.hcbs.mn.qcew <- hcbs.qcew.mn.original %>%
  select(periodyear, naicstitle, areaname, empYear, avg_estabYear, totwageYear, avgwkwageYear) %>%
  rename(year = periodyear,
         planning.region = areaname)
```

<br>

# About the data

The Quarterly Census of Employment and Wages is a census of employers on employment and wages. Data are available by industry, ownership type, states, regions, counties and minor civil divisions.

All establishments covered under the Unemployment Insurance Program are required to report wage and employment statistics quarterly to the Minnesota Department of Employment and Economic Development. Federal government establishments are also covered by the QCEW program. These data are edited and verified for research and analysis purposes.

The dataset below is the employment, establishments, and wages that fall under the NAICS 6216 Code titled "Home Health Care Services". The North American Industry Classification System (NAICS) code 6216 refers to home health care services. This industry includes businesses that provide a range of services, including:

* Skilled nursing services
* Personal care services
* Homemaker and companion services
* Physical therapy
* Medical social services
* Medications
* Medical equipment and supplies
* Counseling
* 24-hour home care
* Occupation and vocational therapy 

Some examples of businesses that fall under the NAICS code 6216 include: Home health care agencies, Visiting nurse associations, Home infusion therapy services, and In-home hospice care services. 

The NAICS code 62 is for the Health Care and Social Assistance sector, which includes businesses that provide health care and social assistance for individuals. 

# Trends in establishments

We will track this using a couple of different methods;

1. Growth/index
2. Share of establishments by location, and
3. Establishments per 65+ population

<br>

## Establishment index{.tabset}

The charts below show that even though Minnesota has experienced 200% growth in the number of home health care establishments since 2006, most of the growth has occurred in the seven county metro. 

By 2023, the Seven County Mpls-St. Paul experienced growth of 230% in this industry, while Northeast had 25% growth. 

### Minnesota

There has been steady growth in the number of home health care businesses in Minnesota since 2005. 

<br>

```{r estab MN, echo=FALSE}
estab.index.mn <- master.hcbs.mn.qcew %>%
  filter(year != max(year)) %>%
  filter(year > 2005) %>%
  mutate(avg_estabYear.index = (avg_estabYear - avg_estabYear[year == min(year)]) / avg_estabYear[year == min(year)],
         data_id = seq(n()))


estab.index.mn.plot <- ggplot(estab.index.mn, aes(year, avg_estabYear.index)) +
  geom_hline(yintercept = 0, color = "black") +
  geom_line(size = 3) +
  geom_point_interactive(size = 2.5, aes(data_id = data_id, tooltip = paste("Minnesota", "\nYear: ", year, "\nNumber of home based health care establishments: ", comma(avg_estabYear), "\nPercent change since ", min(estab.index.mn$year), ": ", percent(avg_estabYear.index, accuracy = .1), sep = ""))) +
  labs(x="", y = "", color="", title = "Change in home based health care establishments since 2005")+
  geom_label_repel(data = filter(estab.index.mn, year == max(year)), aes(label = percent(avg_estabYear.index, accuracy = .1)), show.legend = FALSE, size = 5) +
  scale_y_continuous(labels=scales::percent)+
  scale_x_continuous(breaks = seq(1900, 2050, 2),
                     guide = guide_axis(n.dodge = 2)) +
  theme_line+
  theme(text = element_text(size = 18))

girafe(ggobj = estab.index.mn.plot, width_svg = 7, height_svg = 7) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE),
                 opts_sizing(rescale = FALSE))


```



<br>

## Planning Region

There are significant differences in the number of establishments.  

<br>

```{r estab pr, echo=FALSE}
estab.index.pr <- master.hcbs.pr.qcew %>%
  filter(year != max(year)) %>%
  filter(year > 2005) %>%
  group_by(planning.region) %>%
  mutate(avg_estabYear.index = (avg_estabYear - avg_estabYear[year == min(year)]) / avg_estabYear[year == min(year)]) %>%
  ungroup() %>%
  mutate(data_id = seq(n()))

estab.index.pr.plot <- ggplot(estab.index.pr, aes(year, avg_estabYear.index, color = planning.region)) +
  geom_hline(yintercept = 0, color = "black") +
  geom_line(size = 3) +
  geom_point_interactive(size = 2.5, aes(data_id = data_id, tooltip = paste(planning.region, "\nYear: ", year, "\nNumber of home based health care establishments: ", comma(avg_estabYear), "\nPercent change since ", min(estab.index.mn$year), ": ", percent(avg_estabYear.index, accuracy = .1), sep = ""))) +
  labs(x="", y = "", color="", title = "Change in home based health care establishments since 2005")+
  geom_label_repel(data = filter(estab.index.pr, year == max(year)), aes(label = percent(avg_estabYear.index, accuracy = .1)), show.legend = FALSE, size = 5) +
  scale_y_continuous(labels=scales::percent)+
  scale_x_continuous(breaks = seq(1900, 2050, 2),
                     guide = guide_axis(n.dodge = 2)) +
  scale_color_manual(values = color.pr) +
  theme_line+
  theme(legend.position = "bottom",
        text = element_text(size = 18))

girafe(ggobj = estab.index.pr.plot, width_svg = 7, height_svg = 7) %>%
  girafe_options(opts_selection(type = "none"),
                 opts_toolbar(saveaspng = FALSE),
                 opts_sizing(rescale = FALSE))


```

<br>

## Share of establishments by planning region

Although Minnesota has experienced 200% growth in the industry since 2006, the index charts above indicate that the growth has been uneven. So let's look at the share of establishments by planning region.

<br>

```{r estab share pr, echo=FALSE}
estab.share.pr <- master.hcbs.pr.qcew %>%
  filter(year == max(year)) %>%
  mutate(pct.share = avg_estabYear / sum(avg_estabYear))


```
