---
title: "Summary"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(readxl)
library(janitor)
library(lubridate)
library(systemfonts)
reset_font_cache()
library(ggtext)
```

# Research question

There are five primary drivers in the increase in assisted living and residential care market;

1. Federal and state efforts to move individuals from institutional to noninstitutional or community-based settings.
2. Advances in medical care and technology have altered both demand and need for institutional care.
3. There is a consumer preference shift for more independent living.
4. Assisted living is perceived more positively compared to nursing homes.
5. Assisted living communities are less costly.

<br>

# Exploration of data

Due to these facts, it would seem necessary to highlight the trends in these types of institutions - non-institutional and institutional care;

* Where are they located?
* What is their capacity?


The following is a list of characteristics/attributes that we would like to explore in the data;

Profile of institutional and non-institutional settings for each region of Minnesota

* Nursing homes
  + Private vs. non-profit 
  + Bed availability
  + Reasons for declines in availability - financial vs. workforce shortages
  + Staffing
  + Projected demand

* Assisted living facilities
  + Private vs. non-profit
  + Trends in capacity
  + Services offered
  + Staffing
  
<br>

# Terminology

*Board and care homes*
These small private facilities, also called residential care facilities or group homes, usually have 20 or fewer residents. Rooms may be private or shared. Residents receive personal care and meals, and staff are available around the clock. Nursing and medical care usually are not provided at the home.

In most cases, you must pay the costs of living at a board and care home. Medicare does not cover these costs. Medicaid may provide partial coverage, depending on the state and whether the person is eligible. If the older person has long-term care insurance, check their plan to see if it includes coverage for this type of facility.

*Nursing homes*
Nursing homes, also called skilled nursing facilities, provide a wide range of health and personal care services. Their services focus more on medical care than most assisted living facilities or board and care homes. Services offered in a nursing home typically include nursing care, 24-hour supervision, three meals a day, and assistance with everyday activities. Rehabilitation services, such as physical, occupational, and speech therapy, are also available.

In many cases, people must pay for nursing home care themselves. Medicare generally doesn’t cover long-term stays in a nursing home, but it may pay for some related costs, such as doctor services and medical supplies. Medicaid may also cover some of the costs of nursing homes for people who are eligible based on income and personal resources. If the older person has long-term care insurance, the policy may include some coverage for nursing home care. Check with the insurance company for details.

*Assisted living*
Assisted living is for people who need help with daily care, but not as much help as a nursing home provides. Assisted living facilities range in size from as few as 25 residents to 100 or more. Typically, a few levels of care are offered, and residents pay more if they need extra services or special care.

Assisted living residents usually live in their own apartments or rooms and share common areas. They have access to many services, including up to three meals a day; assistance with personal care; help with medications, housekeeping, and laundry; 24-hour supervision, security, and on-site staff; and social and recreational activities. Some assisted living facilities are part of a larger organization that also offers other levels of care. For example, continuing care retirement communities may also offer independent living and skilled nursing care. Exact arrangements vary by facility and by state.

Most people pay the full costs of assisted living themselves. This option tends to be more expensive than living independently but less expensive than a nursing home. Medicare does not pay for assisted living. Medicaid may provide coverage for some aspects of assisted living, depending on the state and whether the person is eligible. This care option is partially covered by some long-term care insurance policies.

<br>

# Summary

## Nursing Homes

Bed capacity across Minnesota has been declining since 2005. Overall, bed capacity has declined 33% from 2005 to 2024. The most severe declines have occurred in rural counties - Cass (-93%), Grant (-65%), Swift (-69%), Pine (-65%), and Faribault (-63%). By region, the largest declines have occurred in Northwest (-43%), Southwest (-38%), Southeast (33%), Northeast (-31%), Seven County Metro (-30%) and Central (-25%).

When looking at bed capacity as a per capita (age 65+) value, this has increased from having 20 people age 65+ per skilled nursing facility bed in 2005 to 49 people age 65+ per skilled nursing facility bed in Minnesota. The most severe per capita values are located in North Central Minnesota - Pine: 1,387 per bed, Cass: 271 per bed, Hubbarb: 114 per bed, and Crow Wing with 103 per bed. A few of the suburbs also had higher per capita values - Anoka: 148 per bed, Carver: 104 per bed, Dakota: 89 per bed and Washington with 80 per bed.

Interestingly, at a rural-urban basis, the more urban a county group was the higher the per capita of 65+ per bed capacity was - Entirely urban was 60 people per bed and it shrinks to 23 people per bed for our entirely rural county group.

Although it would seem that rural areas might be better off in terms of bed availability, one large concern is the amount of beds that are missing due to facility closures rather than currently operating facilities that have chosen to shrink their bed capacity. A higher percentage of facility closures tracks with how rural a county is - entirely rural counties have lost 26% of their facilities, town/rural mix counties have lost 21%, urban/town/rural county mix has lost 16% while entirely urban counties have lost only 9%.

What this means for rural areas is that there is possibly less flexibility in expanding bed availability if demand increases. For our entirely rural counties, 58% of the decline in bed capacity is from facility closure. 

We know that rural facilities are smaller in size. In 2024, the average facility size was 45 beds compared to 59 for town/rural mix counties, 66 for urban/town/rural mix and 89 beds for entirely urban counties. Even more striking is the difference at the 75% quartile - entirely rural counties had 50 beds where as entirely urban had 105 beds. This shows that there are significantly larger facilities in our more urban areas. 

This is concerning because we see that smaller facilities are significantly more likely to close. The table below shows the statistics and comparitively the facilities that have remained open since 2005 are significantly larger.

## Assisted living

The capacity of assisted living is not nearly as robust in rural Minnesota compared to our more urban areas. Ratios of population 65 and older per assisted living capacity unit shows that they range from 33 people per capacity unit in entirely rural counties down to 15 in entirely urban counties. 

In addition, the ability for assisted living to make up for the decline in nursing home beds is significantly less in rural areas. In our entirely urban counties, assisted living made up over 800% of the loss in nursing home beds. In entirely rural areas, assisted living doesn't make up for the entire loss of nursing home beds - 93%.

<br>

```{r facility size comparison closed vs open}
header <- c("", "Closed", "Open")
row.1 <- c("Q25", 33, 45 )
row.2 <- c("Mean", 59, 72)
row.3 <- c("Median", 48, 62)
row.4 <- c("Q75", 61, 86)

table <- tibble(Statistic = c("Q25", "Mean", "Median", "Q75"), "Closed" = c(33, 59, 48, 61), "Open" = c(45, 72, 62, 86))

kable(table, format = "html", escape = F) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

<br>

## Home health care

Using the QWEC data, there are serious concerns about rural's ability to meet the upcoming demands for elderly care. Admittedly, it's a bit difficult to interpret due to not knowing what locations are being served. It could be that a lot of the metro based home health care establishments are serving rural Minnesota. I sure hope so, because this doesn't look good.

When look at establishments, the seven county metro has 75% of the total while the rest of Minnesota has bout 5% per planning region. This is about the same for employment.

When looking at the population 65 and older per home health care establishment - the seven county metro has 869 individuals per home health care establishment while the other planning regions have 2,500+ per establishment. Employment mimics a similar pattern.

